A non-contextual hidden variable model for QM

### 2016-07-06 finalPaper.tex
made minor changes

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[british,aps,superscriptaddress,nofootinbib,times]{revtex4-1}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm,headheight=2cm,headsep=2cm,footskip=2cm}
\setcounter{secnumdepth}{3}
\usepackage{verbatim}
\usepackage{refstyle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{esint}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
  \theoremstyle{definition}
  \newtheorem{defn}[thm]{\protect\definitionname}
  \theoremstyle{remark}
  \newtheorem{rem}[thm]{\protect\remarkname}
  \theoremstyle{plain}
  \newtheorem*{prop*}{\protect\propositionname}
  \theoremstyle{plain}
  \newtheorem{lem}[thm]{\protect\lemmaname}
  \theoremstyle{plain}
  \newtheorem{prop}[thm]{\protect\propositionname}
  \theoremstyle{definition}
  \newtheorem{example}[thm]{\protect\examplename}

@@

\usepackage{babel}
  \providecommand{\definitionname}{Definition}
  \providecommand{\examplename}{Example}
  \providecommand{\lemmaname}{Lemma}
  \providecommand{\propositionname}{Proposition}
  \providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}

@@

\title{An alternative to contextuality}

@@

\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model, consistent with all predictions of QM (for discrete spaces). We reconsider the no-go theorems on hidden variables and conclude that the notion of contextuality is not a necessary feature of quantum mechanics. Motivated by consistency requirements between Bohmian Mechanics and the said theorems, we provide an alternative view. This is accomplished by identifying a new classical property, we call `multiplicativity'. Advantages of this view are illustrated by considering the implications of non-multiplicativity to non-locality and its relation to contextuality.
\end{abstract}
\maketitle

@@

Background: Since Einstein's \cite{EinsteinEPR} work on completeness
of quantum mechanics and the subsequent seminal work of Bell, assessing
compatibility of hidden variable models has proven to be repeatedly
useful. It has had foundational implications, for instance it gave
a precise meaning to locality, and engendered pragmatic advantages,
such as security, randomness certification etc. While Bell showed
that local hidden variable models can not describe nature, the Kochen
Specker theorem \cite{KochenSpecker} (and related work \cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
ruled out an even larger class of hidden variable models. The notion
of contextuality was identified which also, in recent times has been
harnessed for computation and cryptography. Not all hidden variable
models were, however, proved to be incompatible. Bohm's work \cite{Bohm1,Bohm2}
is one such (in)famous example. This approach has allowed for a precise
interpretation of quantum mechanics.

Previous work {[}the essentials I have already discussed{]}: In this
paper, we take a closer look at the aforesaid no-go theorems and find
that contextuality is in fact not a necessary feature of quantum mechanics.
This has been shown \cite{NoContextuality} by taking disturbances
due to measurements into consideration. Here, however, we construct
a significantly simpler toy model for illustrating not only the aforesaid
but to also delineate a complete alternative to contextuality. The
toy model is a specific case of what we call a c-ingle variable theory.
Its predictions match with those of quantum mechanics for systems
described by arbitrary (discrete) Hilbert spaces.

@@

\begin{defn}
Notation: (a) $\psi\in\mathcal{H}$ would typically represent the
quantum mechanical state of the system (assumed pure), residing in
the Hilbert space $\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean $\mathcal{H}\otimes\mathcal{H}^{\dagger}$, (c) $[\mathcal{H}]$
is defined to mean $(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including hidden variables, (d)
$[\psi]\in[\mathcal{H}]$ will represent the state of the system,
including hidden variables, (e) a prediction map is $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$,
(f) a sequence map is $s:\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}\to[\mathcal{H}]$,
(g) the set of all experimental setups is denoted by $\mathcal{E}$,
(h) a setup map is $e:\hat{\mathcal{H}}\to\mathcal{E}$.
\end{defn}
In addition to the notation, we use the following definition of non-contextuality.
\begin{defn}
A theory is non-contextual, if it provides a map $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$
to explain measurement outcomes. A theory which is not non-contextual
is contextual.\end{defn}
\begin{rem}
Prediction maps are non-contextual.
\end{rem}
In the literature broader definitions have been suggested which declare
a larger set of theories as non-contextual. For our purposes however,
this restricted definition will suffice. The term contextual is used
to suggest that the value an operator takes might depend on which
other compatible observable it is measured with. Since quantum mechanics
(QM) yields deterministic results only in specific situations, it
is imperative to analyse the meaning of contextuality by `completing'
QM. Hidden variable theories do precisely this and we have at our
disposal the heretic theory of Bohm. Circumventing the details of
the theory, we note its salient feature; with each particle, one associates
precisely defined $(q,p)$ and a wavefunction $\psi$. Consequently,
the outcomes of all experiment are predictable, granted the initial
conditions of the system and the apparatus are known (the `hidden
variables'). The essence of Bohmian Mechanics relevant here can be
captured by the following theorem.
\begin{thm}
$\exists$ a (Bohmian) map $m_{B}:\hat{\mathcal{H}},[\mathcal{H}],\mathcal{E}\to\mathbb{R}$,
and a sequence map $s$, s.t. if $m_{B}$ \& $s$ are assumed to describe
the outcomes of measurements \& the resultant state respectively,
then they are consistent with all predictions of QM.
\end{thm}
Observe that given a Stern-Gerlach setup and the initial conditions
(position \& wavefunction of the particle), according to the theorem,
Bohmian mechanics can predict the value of the spin (say $\hat{\sigma}_{z}$)
of the particle. Assume that the state of the particle was $(\left|0\right\rangle +\left|1\right\rangle )/\sqrt{2}$
and that the measurement outcome was spin up. Interestingly, it can
be shown \cite{Detlef} that if the direction of the magnetic field
gradient is flipped while everything else is unchanged, the measurement
outcome will now be spin down. This emphasises the fact that $m_{B}$
is not a prediction map. More importantly it entails that $m_{B}$
is contextual. It must be stressed, however, that the contextuality
in this case is not a statement about the context in terms of compatible
observables. Here no two observables are compatible (excluding identity).

@@

One can show \cite{KochenSpecker} that under reasonable assumptions,
no non-contextual theory can explain all predictions of QM. We re-state
these results and show that nowhere does contextuality appear to be
a logical necessity. Instead we note that non-contextual theories
with certain properties, we call \emph{multiplicativity} are incompatible
with QM. Viewed from this perspective, non-\emph{multiplicativity}
is the non-classical alternative to contextuality in QM. Precise definitions
of this and a related property \emph{sequential multiplicativity},
are given below. 
\begin{defn}
A prediction map $m$ is \emph{multiplicative }iff 
m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi])=f(m(\hat{B}_{1},[\psi]),m(\hat{B}_{2},[\psi]),\dots m(\hat{B}_{N},[\psi])),
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are arbitrary mutually commuting
observables, $f:\hat{\mathcal{H}}^{\otimes N}\to\mathcal{\hat{H}}$
and $[\psi]\in[\mathcal{H}]$. A non-\emph{multiplicative} map is
one that is not multiplicative.
\end{defn}
Note that if $m$ is taken to represent the measurement outcome (in
QM), then for states of the system which are simultaneous eigenkets
of $\hat{B}_{i}$s, $m$ must clearly be multiplicative. It is, however,
not obvious that this property must always hold. To motivate this,
consider the state $\left|11\right\rangle :=\left|1\right\rangle \otimes\left|1\right\rangle $
written in the computational basis, $\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$ and $\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$,
we have $m(\hat{C})=-1$ while $m(\hat{B}_{1})=\pm1$ and $m(\hat{B}_{2})=\pm1$,
independently, according to QM, with probability half. Contrary to
this, from a quick computation, it is clear that if one first measures
$\hat{B}_{1}$ and subsequently measures $\hat{B}_{2}$, then the
product of the results must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define \emph{sequential multiplicativity}
as follows.
\begin{defn}
A prediction map $m$ is \emph{sequentially multiplicative }for a
given sequence map $s$, iff 
m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=f(m(\hat{B}_{1},[\psi_{k_{1}}]),m(\hat{B}_{2},[\psi_{k_{2}}]),\dots,m(\hat{B}_{N},[\psi_{k_{N}}])),
where $\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{(1,2,3\dots N),(2,1,3\dots N)+\text{ permutations, making }N!\text{ terms}\}$,
$\hat{B}_{i}\in\hat{\mathcal{H}}$ are arbitrary mutually commuting
observables, $[\psi_{i}]\in[\mathcal{H}]$, $f:\hat{\mathcal{H}}^{\otimes N}\to\mathcal{\hat{H}}$
and $[\psi_{k+1}]\equiv s(\hat{B}_{k},[\psi_{k}],m(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{1}]$.
\end{defn}
We now discuss the `proofs of contextuality' in the purview of the
aforesaid.
\begin{thm}
Let a map $m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(f(\hat{A}_{1},\hat{A}_{2},\dots))=f(m(\hat{A}_{1}),m(\hat{A}_{2}),\dots)$,
for any arbitrary function $f$, where $\hat{A}_{i}$ are arbitrary
Hermitian operators. If $m$ is assumed to describe the outcomes of
measurements, then no $m$ exists which is consistent with all predictions
of Quantum Mechanics.\end{thm}
Bell construction ($\left|\mathcal{H}\right|\ge4$): Consider an operator
$\hat{B}:=a_{1}\otimes b_{1}+a_{2}\otimes b_{1}+a_{1}\otimes b_{2}-a_{2}\otimes b_{2}$
s.t. $-1\le\hat{a}_{i},\hat{b}_{j}\le1$ $\forall\,i,j\in\{1,2,3,4\}$
(which means that the spectrum of $\hat{a}$ and $\hat{b}$ are bounded
above and below). If we assume that a map $m$ of the aforesaid type
exists, then it can be shown that $\left\langle \hat{B}\right\rangle \le2$.
To see this, consider $\hat{A}_{1}=\hat{a}_{1}\otimes\hat{\mathbb{I}}$
and $\hat{A}_{2}=\hat{\mathbb{I}}\otimes\hat{b}_{1}$, so that if
$\hat{C}:=\hat{A}_{1}\hat{A}_{2}$ then property (2) entails that
$m(\hat{C})=m(\hat{A}_{1})m(\hat{A}_{2})$, viz. $m(\hat{a}_{1}\otimes\hat{b}_{1})=m(\hat{a}_{1}\otimes\hat{\mathbb{I}})m(\hat{\mathbb{I}}\otimes\hat{b}_{1})$.
Note that $m(\hat{*})$, may depend on the state $\left|\psi\right\rangle $
and the hidden variables%
, where $\hat{*}$ is taken to represent $\hat{B}_{k}$s which can
be generalised to all $\hat{a}_{i}$s and $\hat{b}_{j}$s
. Therefore, we average over multiple measurements, which corresponds
to averaging over the hidden variables. This is expressed by $\left\langle m(\hat{a}_{1}\otimes\hat{b}_{1})\right\rangle =\left\langle m(\hat{a}_{1}\otimes\hat{\mathbb{I}})\right\rangle \left\langle m(\hat{\mathbb{I}}\otimes\hat{b}_{1})\right\rangle $.
One can repeat this argument for each term in $\left\langle \hat{B}\right\rangle $.
Using the fact that $-1\le m(\hat{*})\le1$, and that the extrema
will occur only at the extreme values of $m$, one can plugin $\pm1$
for each $\left\langle m(\hat{*})\right\rangle $ to obtain $\left\langle \hat{B}\right\rangle \le2$.
Quantum mechanically, one can show that $\left\langle \hat{B}\right\rangle =2\sqrt{2}$
for the state $\left|\psi\right\rangle =\left(\left|+-\right\rangle -\left|-+\right\rangle \right)/\sqrt{2}$,
where $\hat{\sigma}_{x}\left|\pm\right\rangle =\pm\left|\pm\right\rangle $,
and $\hat{\sigma}_{x,y,z}$ are the Pauli matrices, given in the z-basis
as 
\hat{\sigma}_{x}\doteq\left[\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right],\,\hat{\sigma}_{y}\doteq\left[\begin{array}{cc}
0 & -i\\
i & 0
\end{array}\right],\,\hat{\sigma}_{z}\doteq\left[\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right].
The observables are $\hat{a}_{1}=\hat{\sigma}_{z}$ and $\hat{a}_{2}=\hat{\sigma}_{x}$,
while $\hat{b}_{1}=-\frac{\hat{\sigma}_{z}+\hat{\sigma}_{x}}{\sqrt{2}}$
and $\hat{b}_{2}=\frac{\hat{\sigma}_{z}-\hat{\sigma}_{x}}{\sqrt{2}}$.
Since quantum mechanics violates the inequality, $\left\langle \hat{B}\right\rangle \le2$,
we conclude that the map $m$ can not simultaneously exist and be
consistent with QM.

\begin{proof}
GHZ construction ($\left|\mathcal{H}\right|\ge6$): Consider three
observers with one particle each. Each observer can measure two properties,
call them X and Y, with outcomes $\pm1$. The associated operators
are $\hat{\sigma}_{x}$ and $\hat{\sigma}_{y}$. The state of the
system is $\sqrt{2}\left|\chi_{G}\right\rangle =\left|000\right\rangle -\left|111\right\rangle $
and we also define $\hat{A}:=\hat{\sigma}_{x}\otimes\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$,
$\hat{A}\left|\chi_{G}\right\rangle =\left|\chi_{G}\right\rangle $,
$\hat{B}:=\hat{\sigma}_{y}\otimes\hat{\sigma}_{x}\otimes\hat{\sigma}_{y}$
and $\hat{C}:=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\otimes\hat{\sigma}_{z}$.
A measurement of $\hat{A}$ would correspond to measurement of property
$X,\,Y$ and $Y$ by the three observers respectively. For a map $m$
consistent with measurement outcomes of QM, we must have $m(\hat{A})=m(\hat{B})=m(\hat{C})=+1$.
Property (b) of the map $\implies1=m(\hat{A}\hat{B}\hat{C})=m(\hat{\sigma}_{x}\otimes\hat{\sigma}_{y}\hat{\sigma}_{x}\hat{\sigma}_{y}\otimes\hat{\sigma}_{x})=m(\hat{\sigma}_{x}^{(1)})m(\hat{\sigma}_{y}^{(2)}\hat{\sigma}_{x}^{(2)}\hat{\sigma}_{y}^{(2)})m(\hat{\sigma}_{x}^{(3)})=m(\hat{\sigma}_{x}^{(1)})m(\hat{\sigma}_{x}^{(2)})m(\hat{\sigma}_{x}^{(3)})=m(\hat{D}\equiv\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}),$
where $\hat{\sigma}_{x}^{(1)}\equiv\hat{\sigma}_{x}\otimes\hat{\mathbb{I}}\otimes\hat{\mathbb{I}}$
and so on. It must be stressed that although $\hat{A}$, $\hat{B}$
and $\hat{C}$ mutually commute, $\hat{\sigma}_{2}^{(2)},\hat{\sigma}_{y}^{(2)}$
do not. We used property (b) over both these sets of operators in
the previous step. Returning to the argument, note that $\hat{D}\left|\chi_{G}\right\rangle =-\left|\chi_{G}\right\rangle $,
$\implies m(\hat{D})=-1$ which yields a contradiction. We therefore
conclude that no $m$ with the said properties exists which is consistent
with all predictions of QM.
\end{proof}
Here $m$ maybe viewed as a specific class of prediction maps, that
implicitly depends on the state $[\psi]$. It follows then that the
maps ruled out by the theorem are non-contextual. The set of maps
ruled out, however, can be enlarged by imposing fewer restrictions
on it. This is achieved by the following theorem.
\begin{thm}
Let a map $m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(f(\hat{B}_{1},\hat{B}_{2},\dots))=f(m(\hat{B}_{1}),m(\hat{B}_{2}),\dots)$,
for any arbitrary function $f$, where $\hat{B}_{i}$ are mutually
commuting Hermitian operators. If $m$ is assumed to describe the
outcomes of measurements, then no $m$ exists which is consistent
with all predictions of Quantum Mechanics.\label{thm:KS}\end{thm}
\begin{proof}
Bell construction ($\left|\mathcal{H}\right|\ge4$): The same proof
as the one presented earlier goes through here also since $[A_{1},A_{2}]=0$.
\end{proof}

The following proof is not only more general, but also independent
of the initial state of the system.
\begin{proof}
Peres Mermin ($\left|\mathcal{H}\right|\ge4$):

@@

\begin{rem}
One could in principle assume $m$, to be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(\alpha\hat{B}_{i})=\alpha m(\hat{B}_{i})$, for $\alpha\in\mathbb{R}$,
(c) $m(\hat{B}_{i}^{2})=m(\hat{B}_{i})^{2}$, (d) $m(\hat{B}_{i}+\hat{B}_{j})=m(\hat{B}_{i})+m(\hat{B}_{j})$,
to deduce (d') $m(\hat{B}_{i}\hat{B}_{j})=m(\hat{B}_{i})m(\hat{B}_{j})$
and that $m(\hat{B}_{i})\in$ spectrum of $\hat{B}_{i}$. Effectively
then, condition (b) listed in the theorem is satisfied as a consequence.
Therefore, assuming (a)-(d) as listed here, rules out a larger class
of $m$.
\end{rem}
According to \thmref{KS}, non-contextual maps which are \emph{multiplicative}
must be incompatible with QM. It is apparent that non-contextual maps
which are non-\emph{multiplicative} might be consistent with QM. To
that end, in the following section we first show that this is indeed
the case by removing contextuality from Bohmian Mechanics and supplement
it thereafter by an explicit construction.

>>>>

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[british,aps,superscriptaddress,nofootinbib,times,twocolumn]{revtex4-1}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm,headheight=2cm,headsep=2cm,footskip=2cm}
\setcounter{secnumdepth}{3}
\usepackage{verbatim}
\usepackage{refstyle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{esint}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
  \theoremstyle{definition}
  \newtheorem{defn}[thm]{\protect\definitionname}
  \theoremstyle{remark}
  \newtheorem{rem}[thm]{\protect\remarkname}
  \theoremstyle{plain}
  \newtheorem*{prop*}{\protect\propositionname}
  \theoremstyle{plain}
  \newtheorem{lem}[thm]{\protect\lemmaname}
  \theoremstyle{plain}
  \newtheorem{prop}[thm]{\protect\propositionname}
  \theoremstyle{definition}
  \newtheorem{example}[thm]{\protect\examplename}
  \theoremstyle{definition}
  \newtheorem{notn}[thm]{\protect\notationname}

@@

\usepackage{babel}
  \providecommand{\definitionname}{Definition}
  \providecommand{\examplename}{Example}
  \providecommand{\lemmaname}{Lemma}
  \providecommand{\propositionname}{Proposition}
  \providecommand{\remarkname}{Remark}
  \providecommand{\theoremname}{Theorem}
  \providecommand{\notationname}{Notation}

@@

\title{A non-contextual hidden variable model for QM}

@@

\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model, consistent with all predictions of QM (for discrete spaces). We reconsider the no-go theorems on hidden variables and conclude that the notion of contextuality is not a necessary feature of quantum mechanics. Motivated by consistency requirements between Bohmian Mechanics and the said theorems, we provide an alternative view. This is accomplished by identifying a new classical property, we call `multiplicativity'. Advantages of this view are illustrated by considering the implications of non-multiplicativity to non-locality.%and its relation to contextuality.
\end{abstract}
\maketitle

@@

Background: Einstein's \cite{EinsteinEPR} work on incompleteness of quantum mechanics and the subsequent seminal work of Bell, assessing compatibility of hidden variable models with quantum mechanics has proven to be repeatedly useful. In Bell's case, for instance, it gave a precise meaning to the foundational notion of locality. Further, it has been of pragmatic utility in the areas of security, randomness certification etc. While Bell showed that local hidden variable models can not describe nature, the Kochen Specker theorem \cite{KochenSpecker} (and related work \cite{Gleason,BellOnHiddenVariables,Peres,Mermin}) ruled out an even larger class of hidden variable models. The notion of contextuality was identified which also, in recent times has been harnessed for computation and cryptography [TOOD: add references]. Not all hidden variable models were, however, proved to be incompatible. Bohm's work \cite{Bohm1,Bohm2} is one such (in)famous example. This approach has allowed for an interpretation of quantum mechanics with precise trajectories.

Previous work {[}the essentials I have already discussed{]}: We take a closer look at the aforesaid no-go theorems and find that contextuality is in fact not a necessary feature of quantum mechanics.
This has been shown earlier by taking disturbances due to measurements into consideration \cite{NoContextuality}. In this paper, we construct a significantly simpler toy model for illustrating not only the aforesaid but to also delineate a complete alternative to contextuality. The toy model is a specific case of what we call a c-ingle variable theory. Its predictions match with those of quantum mechanics for systems described by arbitrary (discrete) Hilbert spaces.

@@

\begin{notn} (a) $\psi\in\mathcal{H}$ would typically represent the quantum mechanical state of the system (assumed pure), residing in the Hilbert space $\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined to mean $\mathcal{H}\otimes\mathcal{H}^{\dagger}$ (i.e. the set of Hermitian operators, observables, on the space $\mathcal{H}$), (c) $[\mathcal{H}]$ is defined to mean $(\mathcal{H},\,\mathbb{R}^{\otimes})$, which represents the state of the system including hidden variables, (d) $[\psi]\in[\mathcal{H}]$ will represent the state of the system, including hidden variables, (e) a prediction map is $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$, (f) a sequence map is $s:\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}\to[\mathcal{H}]$, (g) the set of all experimental setups is denoted by $\mathcal{E}$, (h) a setup map is $e:\hat{\mathcal{H}}\to\mathcal{E}$, (i) number of observables is represented by $N \in \mathbb{Z}^+$, (j) $f$ is an arbitrary map from $\hat{\mathcal{H}}^{\otimes N}\to \hat{\mathcal{H}}$ constructed using multiplication and addition of the observables, and multiplication with complex numbers, (k) $\tilde{f}$ is a map from $\mathbb{R}^{\otimes N} \to \mathbb{R}$ constructed by replacing observables in $f$ with real numbers.
\end{notn}

In addition to the notation, we use the following definition of non-contextuality.
\begin{defn}
A theory is non-contextual, if it provides a map $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$ to explain measurement outcomes. A theory which is not non-contextual is contextual.\end{defn}
\begin{rem}
A map is non-contextual if it is of the form $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$; prediction maps are non-contextual.
\end{rem}

In the literature broader definitions have been suggested which declare a larger set of theories as non-contextual. For our purposes however, this restricted definition will suffice. The term contextual is used to suggest that the value an operator takes might depend on which other compatible observable it is measured with. Since quantum mechanics (QM) yields deterministic results only in specific situations, it is imperative to analyse the meaning of contextuality by `completing' QM. Hidden variable theories do precisely this and we have at our disposal the heretic theory of Bohm. Circumventing the details of the theory, we note its salient feature; with each particle, one associates precisely defined $(q,p)$ and a wavefunction $\psi$. Consequently, the outcomes of all experiment are predictable, granted the initial conditions of the system and the apparatus are known (the `hidden variables'). The essence of Bohmian Mechanics relevant here can be captured by the following theorem.

\begin{thm}
$\exists$ a (Bohmian) map $m_{B}:\hat{\mathcal{H}},[\mathcal{H}],\mathcal{E}\to\mathbb{R}$,
and a sequence map $s$, s.t. if $m_{B}$ \& $s$ are assumed to describe
the outcomes of measurements \& the resultant state respectively,
then they are consistent with all predictions of QM.
\end{thm}
Observe that given a Stern-Gerlach setup and the initial conditions (position \& wavefunction of the particle), according to the theorem, Bohmian mechanics can predict the value of the spin (say $\hat{\sigma}_{z}$) of the particle. Assume that the state of the particle was $(\left|0\right\rangle +\left|1\right\rangle )/\sqrt{2}$ and that the measurement outcome was spin up. Interestingly, it can be shown \cite{Detlef} that if the direction of the magnetic field gradient is flipped while everything else is unchanged, the measurement outcome will now be spin down. This emphasises the fact that $m_{B}$ is not a prediction map. More importantly it entails that $m_{B}$ is contextual. It must be stressed, however, that the contextuality in this case is not a statement about the context in terms of compatible observables. Here no two observables are compatible (excluding identity).

@@

One can show \cite{KochenSpecker} that under reasonable assumptions, no non-contextual theory can explain all predictions of QM. We re-state these results and show that nowhere does contextuality appear to be a logical necessity. Instead we note that non-contextual theories with certain properties, we call \emph{multiplicativity} are incompatible with QM. Viewed from this perspective, non-\emph{multiplicativity} is the non-classical alternative to contextuality in QM. Precise definitions of this and a related property \emph{sequential multiplicativity}, are given below.

\begin{defn}
A prediction map $m$ is \emph{multiplicative} iff 
% \[
% \begin{split}
% m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi]) = \\
% f(m(\hat{B}_{1},[\psi]),m(\hat{B}_{2},[\psi]),\dots m(\hat{B}_{N},[\psi])),
% \end{split}
% \]

\begin{align*}
& m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi]) = \\
& \tilde f(m(\hat{B}_{1},[\psi]),m(\hat{B}_{2},[\psi]),\dots m(\hat{B}_{N},[\psi])),
\end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are arbitrary mutually commuting observables and $[\psi]\in[\mathcal{H}]$. A non-\emph{multiplicative} map is one that is not multiplicative.
\end{defn}
Note that if $m$ is taken to represent the measurement outcome (in
QM), then for states of the system which are simultaneous eigenkets
of $\hat{B}_{i}$s, $m$ must clearly be multiplicative. It is, however,
not obvious that this property must always hold. To motivate this,
consider the state of two spin half particles, $\left|11\right\rangle :=\left|1\right\rangle \otimes\left|1\right\rangle $
written in the computational basis, $\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$ and $\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$,
we have $m(\hat{C})=-1$ while $m(\hat{B}_{1})=\pm1$ and $m(\hat{B}_{2})=\pm1$,
independently, according to QM, with probability half. Antithetically it is clear that if one first measures
$\hat{B}_{1}$ and subsequently measures $\hat{B}_{2}$, then the
product of the results must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define \emph{sequential multiplicativity}
as follows.
\begin{defn}
A prediction map $m$ is \emph{sequentially multiplicative} for a
given sequence map $s$, iff 
\begin{align*}
&m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde f(m(\hat{B}_{1},[\psi_{k_{1}}]),m(\hat{B}_{2},[\psi_{k_{2}}]),\dots,m(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*}
where $\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{(1,2,3\dots N),(2,1,3\dots N)$ + permutations, making $N!$ terms$\}$,
$\hat{B}_{i}\in\hat{\mathcal{H}}$ are arbitrary mutually commuting
observables, $[\psi_{i}]\in[\mathcal{H}]$ and $[\psi_{k+1}]\equiv s(\hat{B}_{k},[\psi_{k}],m(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.
\end{defn}
We now discuss the `proofs of contextuality' in the purview of the
aforesaid.
\begin{thm}
Let a map $m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(f(\hat{A}_{1},\hat{A}_{2},\dots))=\tilde f(m(\hat{A}_{1}),m(\hat{A}_{2}),\dots)$,
where $\hat{A}_{i}$ are arbitrary Hermitian operators. If $m$ is assumed to describe the outcomes of measurements, then no $m$ exists which is consistent with all predictions of Quantum Mechanics.\label{thm:det} \end{thm}
Here $m$ maybe viewed as a specific class of prediction maps, that
implicitly depends on the state $[\psi]$. It follows then that the
maps ruled out by the theorem are non-contextual. The set of maps
ruled out, however, can be enlarged by imposing fewer restrictions
on it. This is achieved by the following theorem.

\begin{thm}
Let a map $m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde f(m(\hat{B}_{1}),m(\hat{B}_{2}),\dots)$,
for any arbitrary function $f$, where $\hat{B}_{i}$ are mutually
commuting Hermitian operators. If $m$ is assumed to describe the
outcomes of measurements, then no $m$ exists which is consistent
with all predictions of Quantum Mechanics.\label{thm:KS}\end{thm}

% The following proof is not only more general, but also independent
% of the initial state of the system.
\begin{proof}
Peres Mermin ($\left|\mathcal{H}\right|\ge4$):

@@

\begin{rem}
One could in principle assume $m$, to be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(\alpha\hat{B}_{i})=\alpha m(\hat{B}_{i})$, for $\alpha\in\mathbb{R}$,
(c) $m(\hat{B}_{i}^{2})=m(\hat{B}_{i})^{2}$, (d) $m(\hat{B}_{i}+\hat{B}_{j})=m(\hat{B}_{i})+m(\hat{B}_{j})$,
to deduce (d') $m(\hat{B}_{i}\hat{B}_{j})=m(\hat{B}_{i})m(\hat{B}_{j})$
and that $m(\hat{B}_{i})\in$ spectrum of $\hat{B}_{i}$. Effectively
then, condition (b) listed in the theorem is satisfied as a consequence.
Therefore, assuming (a)-(d) as listed here, rules out a larger class
of $m$.
\end{rem}
According to \thmref{KS}, non-contextual maps which are \emph{multiplicative}
must be incompatible with QM. It is apparent that non-contextual maps
which are non-\emph{multiplicative} might be consistent with QM. To
that end, in the following section we first show that this is indeed
the case by constructing an explicit model and supplement it by removing contextuality from Bohmian Mechanics.
